---
title: "Integrated KSAO Framework for Substance Use Disorder Counselors"
subtitle: "A Comprehensive Competency Map Derived from CASAC Textbook Analysis"
author: "KSAO Workforce Development Project"
date: today
format: 
  html:
    toc: true
    number-sections: true
    theme: cosmo
    code-fold: true
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
execute:
  echo: false
---

```{r setup}
#| include: false
library(readr)
library(knitr)
library(dplyr)
```

# Introduction

This document presents the integrated Knowledge, Skills, Abilities, and Other Characteristics (KSAOs) framework for substance use disorder counselors, derived from a comprehensive analysis of the CASAC (Credentialed Alcoholism and Substance Abuse Counselor) certification textbook. The analysis focused on Chapters 1-2 and Appendices, using Google's Gemini 2.5 Pro model to identify and categorize competencies.

This framework is designed to serve as a foundation for evidence-based workforce development in addiction counseling, moving beyond traditional certification requirements to focus on actual competencies required for effective practice.

# Methodology

## Data Sources

The analysis was conducted on the following chapters from the CASAC textbook:

1. **Chapter 1**: Covering SUD terminology, neurobiology, models of addiction, risk factors, stages of addiction, common substances, recovery principles, and the Stages of Change model.

2. **Chapter 2**: Focusing on systems of care, telehealth, harm reduction, trauma-informed systems, social determinants of health, and related topics.

3. **Appendices**: Including commonly used drugs, mental health conditions, medications, and IC&RC domains and job tasks.

## Analysis Process

The analysis followed a rigorous five-step process:

1. **OCR Text Extraction**: Used Gemini 2.5 Pro's multimodal capabilities to extract text from TIFF images of textbook pages.

2. **Chapter-Level KSAO Analysis**: Each chapter was analyzed individually to identify KSAOs, with documentation of the thinking process.

3. **KSAO Integration**: Individual chapter analyses were integrated into a comprehensive framework, resolving redundancies and inconsistencies.

For each KSAO, the analysis identified:
- Clear name/title
- Complete description
- Classification (Knowledge, Skill, Ability, or Other characteristic)
- Specificity level (general or specialized)
- Related O*NET occupational categories
- Stability/malleability (fixed vs. developable)
- Explicit/tacit orientation (explicitly taught vs. tacitly acquired)
- Prerequisites or developmental relationships

# Integrated KSAO Framework

```{r}
#| include: false
# Read the integrated framework file
framework_file <- "../output/full_analysis/integrated_ksao_framework.txt"
framework_text <- readLines(framework_file)

# Find where the actual framework starts (after the thinking process)
start_line <- which(grepl("^# INTEGRATED KSAO FRAMEWORK", framework_text, ignore.case = TRUE))
if(length(start_line) == 0) {
  # If exact heading not found, look for any heading that might indicate the framework
  start_line <- which(grepl("^#.*KSAO.*FRAMEWORK", framework_text, ignore.case = TRUE))
}

if(length(start_line) > 0) {
  framework_content <- framework_text[start_line:length(framework_text)]
} else {
  framework_content <- framework_text
}
```

```{r, results='asis'}
# Output the framework content
cat(paste(framework_content, collapse = "\n"))
```

# Hierarchical Structure and Developmental Relationships

This section presents the hierarchical organization of KSAOs, showing how they relate to each other in terms of:

1. **Dimensions vs. Sub-dimensions**: Major competency domains and their components
2. **Developmental Sequences**: The logical progression for acquiring competencies
3. **Foundational Relationships**: Which KSAOs serve as prerequisites for others

```{r}
#| include: false
# Find where the hierarchical structure section starts
hierarchy_start <- which(grepl("^# Hierarchical Structure", framework_text, ignore.case = TRUE))
if(length(hierarchy_start) == 0) {
  hierarchy_start <- which(grepl("^#.*Hierarchical|^#.*Structure|^#.*Relationships", framework_text, ignore.case = TRUE))
}

if(length(hierarchy_start) > 0 && hierarchy_start < start_line) {
  # If hierarchy section comes before the framework section
  hierarchy_end <- start_line - 1
  hierarchy_content <- framework_text[hierarchy_start:hierarchy_end]
} else if(length(hierarchy_start) > 0) {
  # If hierarchy section comes after the framework section
  hierarchy_content <- framework_text[hierarchy_start:length(framework_text)]
} else {
  hierarchy_content <- character(0)
}
```

```{r, results='asis'}
# Output the hierarchy content if it exists
if(length(hierarchy_content) > 0) {
  cat(paste(hierarchy_content, collapse = "\n"))
}
```

# Applications for Workforce Development

The integrated KSAO framework presented in this document can be applied in multiple ways to enhance workforce development in the field of substance use disorder counseling:

1. **Curriculum Design**: Educational institutions can use this framework to design comprehensive training programs that systematically address all required competencies.

2. **Certification Standards**: Certification bodies can align requirements with evidence-based KSAOs rather than solely focusing on knowledge or hours of training.

3. **Professional Development**: Practitioners can identify gaps in their skill set and target specific areas for improvement.

4. **Supervision Focus**: Clinical supervisors can use the framework to provide structured guidance for developing counselors.

5. **Assessment Tools**: The framework provides a foundation for developing assessment instruments to measure counselor competency.

# Conclusion

This integrated KSAO framework represents a significant advancement in understanding the competencies required for effective substance use disorder counseling. By systematically extracting and organizing KSAOs from authoritative textbook material, this project demonstrates how AI-assisted analysis can support evidence-based workforce development.

The framework highlights the complex, multifaceted nature of SUD counseling, requiring a diverse set of knowledge areas, clinical and interpersonal skills, cognitive abilities, and professional characteristics. It also reveals the developmental nature of these competencies, with clear progression pathways and foundational relationships.

Future work should extend this analysis to the remaining textbook chapters to create a truly comprehensive framework, followed by validation studies with subject matter experts and practicing counselors.